### 2016-01-28 paper/response.tex

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color} 
\usepackage[usenames,dvipsnames,table]{xcolor}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
%\doublespacing

@@

\usepackage{multirow}

\usepackage{color}
\usepackage{csquotes}
\usepackage{ulem}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

\definecolor{dkred}{rgb}{0.75,0,0}
\definecolor{dkgreen}{rgb}{0,0.5,0}
\definecolor{dkblue}{rgb}{0,0,0.75}
\definecolor{dkpurple}{rgb}{.375,0,.375}
\definecolor{gray}{rgb}{0.5,0.5,0.5}

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
%\newcommand{\removed}[1]{\textcolor{dkred}{#1}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}

\newenvironment{response}{\fontfamily{cms}\selectfont\small}{\par}

\renewcommand{\rmdefault}{cmr}
\renewcommand{\sfdefault}{lmss}

@@

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung (Response to Reviewers)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin, 
Stephanie Forrest, 
Soumya Banerjee,
Candice Clay,
Judy Cannon, 
Melanie Moses, 
Frederick Koster
\end{flushleft}
\vspace{1cm}

We thank the reviewers for their constructive and insightful comments.  We have done our best to address each point raised and feel the incorporation of the reviewers' comments has greatly improved the paper.  We include detailed responses to each comment below.

@@

\section*{Reviewer 1}

\begin{enumerate}

\item \textbf{To me, the most interesting result would be estimating parameters for the dynamics of the virus and production of chemokines but the authors provide little to no details of how this was done. This is not sufficient as the model has so many parameters I would like to see how you estimated those. What are the model fits? Which parameters were fixed in fitting and why? What if Mitchell et al. study has parameters which are not compartible with your experiments? you need to explore this much deeper than the current ms. does.}

\begin{response}
We added one new paragraph and modified another in the Models for Parameter Estimation subsection to address the reviewer's three main questions, as well as others the reviewer did not explicitly ask for.  In summary:
\begin{itemize}
\item The model was fit using a genetic algorithm to minimize the sum of the squared log residuals between the model and the data.  In an earlier analysis the GA fits were superior to non-linear regression methods in both the Berkeley Madonna and Matlab software suites.
\item All parameters were held constant during the fitting procedure aside from the chemokine secretion rate.  
\item The model was fit to experimental data measured during the experiments reported in Mitchell 2011.  However, the chemokine data (Fig. 3, Table S1) have not been published previously.
\item 1,000 bootstrapping runs were performed to generate confidence intervals over each estimated parameter value.
\end{itemize}

The new text to address the comment is:

\begin{displayquote}
\new{To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig. 3, Table S1).  The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus (Mitchell et al., 2011).  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..}

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in (Mitchell et al., 2011) Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.   \new{Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.  Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.}  The results of these fits are shown in Table 1 \new{(viral secretion rates are from previous fits in Mitchell et al., 2011)}.  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2). \\

... Eq. 1 ... \\

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.  \removed{Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.  The results of these fits are shown in Table 1.  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).}
\end{displayquote}
\end{response}

\item \textbf{I am not completely clear on many of the parameter choices in your model. Yes, you cite several studies for the parameter choice but I would like to see a more thorough review of which parameters you chose and why. The issue is that your model is not ODE model used in many of the studies you cite, so it is not straightforward to go from one model to another assuming that parameters must be the same (or similar). Specifically, how do you obtain 1257 cell/hour emergence from LNs? T cells decay with average lifespan of 3 days? IgM-mediated viral clearance increase of 10 fold? Constant T cell production after day 5?}

\begin{response}
We agree with the reviewer that there is no easy answer for which parameters to include or exclude from the model.  We added a new paragraph to the beginning of the Model Parameters subsection of the Modeling and Methods section to better explain our reasoning:

\begin{displayquote}
\new{We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection. Thus, we included parameters that affected virus, T cells, and chemokines. With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.  The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.  Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.  We added IgM because many studies have argued for the importance of antibody mediated virus clearance. However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.}
\end{displayquote}

The reviewer is correct in observing that agent-based modeling is different from ODE models, we think that using parameters from the ODE models is a good first estimate.  We included parameters that ODE models have determined (cited in Table 2) and in cases where appropriate (such as the viral infection rate), we adapted values from an ODE model to mirror the expected behavior in a spatial environment.

The specific parameters identified by Reviewer 1, each one is listed with a citation or justification in Table 2.  For the case of the T cell production rate, there is also a subsection in the paper that describes our methodology (Models for Parameter Estimation).  For the case of the T cell decay rate, three days was an error as the correct value was four days.  This has been fixed in the manuscript.  These values are first referenced in the Model Implementation subsection of the Materials and Methods section, before the justification for the values are referenced.  To address this inconsistency, we moved the Model Parameters subsection to appear \textit{before} the Model Implementation subsection, and we added more direct references to Table 2 and the Models for Parameter Estimation subsection for clarity.

The main finding of our study through sensitivity analysis shows that few parameters actually affect influenza clearance. Our findings confirm that despite the difficulty selecting parameters to populate the model, selection of most parameters is not likely to be a crucial factor in determining our results.​

\end{response}

\item \textbf{The number of simulations performed seems small (50). Will statistics be better if you run more reasonable number of simulations as recommended for bootstrap analyses (e.g., 500-1000).}

\begin{response}
The spatial nature of the model made it difficult to perform a full 1,000 sample bootstrapping run. Further, bootstrapping works by shuffling residuals of a model fit to generate new data sets for the purpose of generating confidence intervals over fit parameters.  In this case, we were simply interested in sampling the variance of the stochastic model itself.  We chose 50 runs of each model parameter set as that number was at the larger end of what was computationally tractable.

To satisfy the reviewer's request, we performed an additional 50 samples of each modeled strain of influenza, doubling our sample size.  Avian influenza's maximum standard deviation changed from 37.2 to 35 (5.9\%), seasonal influenza changed from 242 to 212 (12.4\%), and pandemic influenza changed from 230 to 231 (0.4\%).  The text and Figure 4 have been updated to include the new model runs.

We think the small change between the old and new values demonstrates that 100 runs is sufficient, but we can perform more samples if the reviewer deems it necessary.
\end{response}

\item \textbf{How important is the assumption of diffusion rate of chemonkines being slower than rate of viral spread? What if chemokine spreads faster? Will control be better?}

\begin{response}
This is a good question.  It is important to note that this is a different question than that regarding the direct comparison between the viral diffusion rate and the chemokine diffusion rate as the chemokine's effect is only seen in the gradient of the concentration and the location of the local maxima, while the viral effect is seen in the spread and concentration of the virions themselves.

We added an explanatory text to the Spatial Effects subsection of the Results section to better address this question:

\begin{displayquote}
Finally, the spatial nature of our model reveals that chemokines can diffuse more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.  It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.  Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.  \new{The differences in the infection outcomes between the three strains illustrate the effect of the infection spreading faster than the chemokine maxima can move in the case of pH1N1.  The velocity of the chemokine maxima positions depends on a combination of many factors, including the chemokine diffusion, decay, delay, and secretion rates (Table 2).}
\removed{This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.}
\end{displayquote}
\end{response}

\item \textbf{Your references are not formatted properly. Please correct.}

\begin{response}
We use the Mendeley reference manager to manage our citations.  We used the official JTB style formatter for Mendeley, along with the specific bibliography style file supplied by JTB, \texttt{elsarticle-num-names.bst}, to format our references with latex and bibtex.  We are happy to continue working with the editor to make our reference style compatible with the expectations of the journal but require further guidance on this point.
\end{response}

@@

\end{enumerate}

\section*{Reviewer 2}

\begin{itemize}

%\item \textbf{(Introduction) P5 L3: delete one "the"}
%
%\begin{response}
%Done
%\end{response}

\item \textbf{(Introduction) P5 L4: ``Pathogenicity is a function of the virus"....too vague. Function of what in the virus?}

\begin{response}
We added a new sentence to clarify our goals.  The question, ``a function of what in the virus?" is exactly the question this paper tries to answer.

\begin{displayquote}
Pathogenicity is a function of \new{both} the virus and its interaction with the immune response. \new{Our model explores how various features of the virus and the host immune system interact to produce observed differences between strains.}
\end{displayquote}

\end{response}

\item \textbf{P8: ``T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds. If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds. If a circulating cell encounters chemokine, it changes state to chemotaxing and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days." Add some references for all these data.}

\begin{response}
This statement is related to Reviewer 1, point 2 (see above for how we addressed it). In summary, we added a new section describing our choice of the parameter set.  We moved the Model Parameters subsection up to appear before the Model Implementation subsection.  We also added references to Table 2, which contains citations and justifications for each of our parameter values.
\end{response}

\item \textbf{P8: ``IgM is modeled by...". IgM never defined.}

\begin{response}
We have added a new sentence to the Model Definition section.

\begin{displayquote}
\new{After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.}
\end{displayquote}
\end{response}

\item \textbf{P11: ``This ODE was fit to data taken from (Miao et al., 2010) using Matlab's nlinfit function in order to obtain a value for $<$sigma$>$. The final value was found to be 1,257 per hour." Any confidence intervals over sigma?}

\begin{response}
Although we don't report confidence intervals, we explain how we obtained the value in the section ``Models for Parameter Estimation".  $\sigma$ is part of our sensitivity analysis (T cell secretion rate) and shows its value does not have a strong effect on the model results within a biologically plausible range.
\end{response}

\item \textbf{P12: ``Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig. 4). " So 50 runs with the same parameter set? Not clear}

\begin{response}
We have expanded our sample from 50 to 100 runs as per Reviewer 1, point 3.  We have clarified that all 100 runs use the same parameter set.
\begin{displayquote}
Because the CyCells model is stochastic, we conducted \new{100}\removed{50} runs of the model \new{with the same parameter set} for each of the three influenza strains, each run initialized with a unique random seed (Fig. 4).  
\end{displayquote}
\end{response}

\item \textbf{P12: ``We estimated variance across runs by calculating the standard deviation at each time point." Standard deviation of what? Viral load? Please specify.}

\begin{response}
We added a sentence to the beginning paragraph of that (Model Results) section:
\begin{displayquote}
Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig. 4).  \new{We count the number of infected cells in the simulated plaque at every time point.}  For the avian and seasonal strains, all model runs cleared the infection by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection. \\

We estimated variance across runs by calculating the standard deviation \new{of the number of infected cells} at each time point. ...
\end{displayquote}
\end{response}

\item \textbf{P13: Not clear how the Sensitivity Analysis has been performed. Were the parameters varied simultaneously? It appears that they have been varied independently (``Each of the remaining sixteen parameters was varied independently"). See PMCID: PMC2570191 for a comprehensive review of global sensitivity analysis methods, especially in the context of an ABM/stochastic model.}

\begin{response}

We thank the reviewer for this suggestion and reference. We performed a new sensitivity analysis using Latin hypercube sampling (LHS) and partial rank correlation coefficient (PRCC) analysis following the suggested citation to supplement our original analysis.  The new sensitivity analysis was performed only for the sH1N1 data due to computational constraints.  333 sample points involving 17 parameters (those of Table 3 plus the viral secretion rate) were generated using LHS and each point was evaluated three times, for a total of 999 model runs.  One sample point caused undefined model behavior and was removed, leaving 996 samples.

The results of the PRCC analysis confirm and quantitate the original sensitivity analysis.  As we found previously, all sensitive parameters remain sensitive under PRCC ($P < 0.01$ with a high absolute Spearman's $\rho$).  Infected Cell expression time and T Cell Expected Lifespan at the FOI were confirmed to be significant with lower values of $\rho$.  These results have been added as a new paragraph to the Sensitivity Analysis subsection of the Results section in the main paper (text), Table 3 in the main paper (below), and expanded upon in the Supplement with text (S2.4) and three new figures (Figs. S6-S8).

\begin{displayquote}
\new{Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).  The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).  This indicates our less computationally intensive OFAT analysis is accurate.}
\end{displayquote}

%\newpage

\begin{displayquote}

### 2016-01-28 paper/Levin_JTB_Clean.tex
In process of drafting Response 3

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

@@

\usepackage{color}
\usepackage[usenames,dvipsnames,table]{xcolor}
%\usepackage{ulem}

@@

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we develop an agent based spatial model to focus on T cell migration from lymph nodes through the vascular system to sites of infection. We use our model to investigate how different strains of influenza modulate this process.

We calibrate the model using viral and chemokine secretion rates we measure \textit{in vitro} together with values taken from literature.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.  % -- Fred --
%We find that as plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.  Further, differences in the diffusion rates of chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  While our sensitivity analysis identifies viral parameters as the most significant factor in determining the course of the infection, we are also able to determine an upper bound of T cell search time using the model.  Finally, we use the model parameters to determine an upper bound on T cell search time in the model.  Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.
% -- Fred , after p2s2 -- 
In this model comparing three influenza viruses, plaque expansion is governed primarily by the replication rate of the virus strain, and the efficiency of the T cell search-and-kill is limited by the density of infected epithelial cells in each plaque.  Thus for each virus there is a different threshold of T cell search time after which recruited T cells are unable to control further expansion.  Future models could use this relationship to more accurately predict control of the infection.

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \citep{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.  Pathogenicity is a function of the virus and its interaction with the host immune response.

Identifying the critical components of this complex process, and how they interact, is key to understanding viral pathogenicity and designing therapeutic interventions.  We address this challenge through computational modeling, which allows us to study the relative contributions of different aspects of the host immune response and to account for strain-specific viral dynamics in the model.  Several earlier theoretical models have been proposed to study specific aspects of influenza infections. Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but ignore localized spatial effects.  Spatial models, on the other hand, have been developed to examine interactions between dendritic cells and T cells in the lymph node \citep{Beauchemin2005, Beltman2007, Zheng2008, Mirsky2011, Celli2012, Vroomans2012, Textor2014}, but have not been extended to consider local conditions in the lung.   In this paper, we examine how the unique properties of three different strains of influenza, in the presence of chemokines, affect T-cell search in the human lung.

@@

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay. \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die. \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour. \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour. T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of three days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.

In addition to T cells, the model contains two types of particles: virus and chemokine. Both are produced at constant rates by expressing epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell. Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.

@@

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.  Initial population sizes and parameter values are taken from the previous study.
{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{Eq. 1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}

$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.  Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.  The results of these fits are discussed in the main paper and shown in Table 1.  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

@@

%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).  The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \citep{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..

%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref. \citep{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).  Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There was no positive correlation between viral production and induced chemokine production across the three strains.

@@

Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).  For the avian and seasonal strains, all model runs cleared the virus by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.

@@

We estimated variance across runs by calculating the standard deviation at each time point.  For aH5N1, the maximum standard deviation was 37, which occurred at day 5.05 where the mean value was 259.  For sH1N1, the maximum standard deviation was 242, which occurred at day 4.93 where the mean value was 2,607.  For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018 (Fig.~\ref{fig:variance}).

>>>>

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
%\usepackage{amssymb}

@@

\usepackage{color}
\usepackage[usenames,dvipsnames,table]{xcolor}
\usepackage{ulem}

%\usepackage[normalem]{ulem}

@@

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}
\newcommand{\drew}[1]{{\color{dkgreen}#1}}
\newcommand{\fred}[1]{{\color{dkblue}#1}}
\newcommand{\steph}[1]{{\color{dkpurple}#1}}

@@

Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.  Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.  One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.  Here, we develop an agent based spatial model to focus on T cell migration from lymph nodes through the vascular system to sites of infection. We use our model to investigate whether different strains of influenza modulate this process.

We calibrate the model using viral and chemokine secretion rates we measure \textit{in vitro} together with values taken from literature.  The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.  % -- Fred --
%We find that as plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.  Further, differences in the diffusion rates of chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.  While our sensitivity analysis identifies viral parameters as the most significant factor in determining the course of the infection, we are also able to determine an upper bound of T cell search time using the model.  Finally, we use the model parameters to determine an upper bound on T cell search time in the model.  Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.
% -- Fred , after p2s2 -- 
In this model comparing three influenza viruses, plaque expansion is governed primarily by the replication rate of the virus strain, and the efficiency of the T cell search-and-kill is limited by the density of infected epithelial cells in each plaque.  Thus for each virus there is a different threshold of T cell search time above which recruited T cells are unable to control further expansion.  Future models could use this relationship to more accurately predict control of the infection.

@@

Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \citep{Who2009}.  Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the mechanisms that control this variability are not well understood.  Pathogenicity is a function of both the virus and its interaction with the immune response. Our model explores how various features of the virus and the host immune system interact to produce observed differences between strains.

Identifying the critical components of this complex process, and how they interact, is key to understanding viral pathogenicity and designing therapeutic interventions.  We address this challenge through computational modeling, which allows us to study the relative contributions of different aspects of the host immune response and to account for strain-specific viral dynamics in the model.  Several earlier theoretical models have been proposed to study specific aspects of influenza infections. Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but are unable to perceive localized spatial effects.  Spatial models, on the other hand, have been developed to examine interactions between dendritic cells and T cells in the lymph node \citep{Beauchemin2005, Beltman2007, Zheng2008, Mirsky2011, Celli2012, Vroomans2012, Textor2014}, but have not been extended to consider local conditions in the lung.   In this paper, we examine how the unique properties of three different strains of influenza, in the presence of chemokines, affect T-cell search in the human lung.

@@

The model is initialized with a single infected cell.  After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).  Virus from expressing cells diffuses locally, infecting neighboring cells. Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI. After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.  After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.  These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.

We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection. Thus, we included parameters that affected virus, T cells, and chemokines. With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.  The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.  Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.  We added IgM because many studies have argued for the importance of antibody mediated virus clearance. However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.

@@

In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}. \texttt{Healthy} cells remain unchanged unless infected by virus. Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay (Table \ref{tab:parameters}). \texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die (Table \ref{tab:parameters}). \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells. \texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour (Table \ref{tab:parameters}). \texttt{Dead} cells take up space and do not regenerate over the course of an infection.  Cell regrowth is not implemented as the rate of regrowth is unknown.

T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.    T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour (Models for Parameter Estimation, Table \ref{tab:parameters}). T cell travel time from the lymph node to a random location on the lung's surface is six seconds.  If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.  If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI. Circulating T cells decay exponentially with an average lifespan of four days.  Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells. Chemotaxing T cells decay exponentially with an average lifespan of two hours.  Justifications for these parameter values are listed in Table \ref{tab:parameters}.

In addition to T cells, the model contains two types of particles: virus and chemokine. Both are produced at constant rates by expressing epithelial cells.  Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell. Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells. Both particle types decay exponentially.  IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day (Table \ref{tab:parameters}).

@@

To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig. \ref{fig:data}, Table S1).  The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus \citep{Mitchell2011}.  IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..

We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq. 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.   Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.  Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.  The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).  IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).

{\footnotesize
\begin{equation*}
\begin{aligned}
\dot{T} &= - \beta T V \\
\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
\dot{F} &=  I_{1 \tau_2} \\
\dot{C} &= r I_{1 \tau_3} - d C \\
\end{aligned}
\tag{Eq. 1}
\label{eq:dde}
\end{equation*}
\vspace{.05in}
}
$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.

@@

% Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).  The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).  Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.  There was no positive correlation between viral production and induced chemokine production across the three strains.

@@

Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig. 4).  We count the number of infected cells in the simulated plaque at every time point.  For the avian and seasonal strains, all model runs cleared the infection by day 10.  Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.

@@

We estimated variance across runs by calculating the standard deviation at each time point.  For aH5N1, the maximum standard deviation was 35, which occurred at day 4.8 where the mean value was 257.  For sH1N1, the maximum standard deviation was 212, which occurred at day 5.1 where the mean value was 2,528.  For pH1N1, the maximum standard deviation was 231, which occurred at day 5.1 where the mean value was 6,988 (Fig.~\ref{fig:variance}).
